---
title: "48hSILI_vs_top10dilli"
output: html_notebook
---

## Set directories
```{r}
# input directory
if(!dir.exists("INPUT")){
  dir.create(path = file.path(getwd(), "INPUT"))
}
input_dir <- file.path(getwd(), "INPUT")

# output directory
if(!dir.exists("OUTPUT")){
  dir.create(path = file.path(getwd(), "OUTPUT"))
}
output_dir <- file.path(getwd(), "OUTPUT")

# Load required packages
library(ggplot2)
library(ggrepel)
library(dplyr)


```


## load the eg scores of SILI and DILI data
```{r}
library(dplyr)
library(ggplot2)
eg_scores_sili <- read.delim(file.path(output_dir,"UploadFile_moduleTable-2024-06-17.txt"))

eg_scores_dili <- readRDS(file.path(output_dir,"eg_score.rds"))

eg_scores_sili_filtered <- eg_scores_sili %>%
  filter(time == "2")
```


##select the top 10 compounds with the highest pearR score for mouse
```{r}
eg_scores_dili_filtered1 <- eg_scores_dili %>%
  filter(experiment == "PHALLOIDIN", conc == "1", time == "1") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))
eg_scores_dili_filtered2 <- eg_scores_dili %>%
filter(experiment == "MONOCROTALINE", conc == "10", time== "8") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered3 <- eg_scores_dili %>%
filter(experiment == "DANAZOL", conc == "1000", time== "1") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered4 <- eg_scores_dili %>%
filter(experiment == "METHYLENE DIANILINE", conc == "100", time== "4") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered5 <- eg_scores_dili %>%
filter(experiment == "TRIMETHADIONE", conc == "500", time== "1") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered6 <- eg_scores_dili %>%
filter(experiment == "THIOACETAMIDE", conc == "45", time== "4") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered7 <- eg_scores_dili %>%
filter(experiment == "THIOACETAMIDE", conc == "45", time== "15") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))%>%
  mutate(experiment_dili = paste0(experiment_dili, "2"))

eg_scores_dili_filtered8 <- eg_scores_dili %>%
filter(experiment == "CARBON TETRACHLORIDE", conc == "300", time== "15") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered9 <- eg_scores_dili %>%
filter(experiment == "THIOACETAMIDE", conc == "45", time== "29") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))%>%
  mutate(experiment_dili = paste0(experiment_dili, "3"))

eg_scores_dili_filtered10 <- eg_scores_dili %>%
filter(experiment == "MONOCROTALINE", conc == "10", time== "15") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))%>%
  mutate(experiment_dili = paste0(experiment_dili, "2"))

eg_scores_dili_sili_filtered <- rbind(
  eg_scores_dili_filtered1, eg_scores_dili_filtered2, eg_scores_dili_filtered3) 
  # eg_scores_dili_filtered4, eg_scores_dili_filtered5,
  # eg_scores_dili_filtered6, eg_scores_dili_filtered7, eg_scores_dili_filtered8)
  # eg_scores_dili_filtered9, eg_scores_dili_filtered10)
```

```{r}
eg_scores_dili_sili_filtered <- eg_scores_dili_sili_filtered %>%
  mutate(experiment_sili = case_when(
    experiment_sili == "uploadFile1_NCSUPHxcaudate" ~ "rat 48h",
    experiment_sili == "uploadFile2_liverhepatocytesPHx48hr" ~ "mouse 48h",
    TRUE ~ experiment_sili  # Keep other values unchanged
  ))

```


##Create correlation plots for the mouse 48h model
```{r}

# Initialize plot list
plot_list <- list() 

# Loop through each experiment and create a correlation plot
unique_experiments <- unique(eg_scores_dili_sili_filtered$experiment_dili)

for (exp in unique_experiments) {
  # Subset the data for the current experiment
  subset_data <- eg_scores_dili_sili_filtered %>% 
    filter(experiment_dili == exp)%>%
    filter(experiment_sili == "mouse 48h")
  
  # Check if subset_data is non-empty
  if (nrow(subset_data) > 0) {
   
    # Identify the top 5 and bottom 5 modules based on eg_score
    top_5_dili <- subset_data %>% top_n(5, eg_score_dili)
    top_5_sili<- subset_data %>% top_n(5, eg_score_sili)
    bottom_5_dili <- subset_data %>% top_n(-5, eg_score_dili)
    bottom_5_sili <-subset_data %>% top_n(-5, eg_score_sili)
  
   # Label the top and bottom modules
    label_data1 <- bind_rows(top_5_dili, bottom_5_dili) %>%
      mutate(experiment_sili = case_when(
        experiment_sili == "uploadFile2_liverhepatocytesPHx48hr" ~ "rat_dataset",
        experiment_sili == "uploadFile1_NCSUPHxcaudate" ~ "mouse_dataset",
        TRUE ~ as.character(experiment_sili)
      ))

    
    label_data2 <- bind_rows(bottom_5_sili, top_5_sili)%>%
      mutate(experiment_sili = case_when(
        experiment_sili == "uploadFile2_liverhepatocytesPHx48hr" ~ "rat_dataset",
        experiment_sili == "uploadFile1_NCSUPHxcaudate" ~ "mouse_dataset",
        TRUE ~ as.character(experiment_sili)
      ))

    
    label_data <- rbind(label_data2, label_data1)
    
    p <- ggplot(subset_data, aes(x = eg_score_sili, y = eg_score_dili, color = experiment_sili)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  geom_label_repel(data = label_data1, aes(label = module_number), 
                   box.padding =0.1, color = "red", segment.color = "grey50") +
  geom_label_repel(data = label_data2, aes(label = module_number), 
                   box.padding = 0.8, color = "blue", segment.color = "grey50") +
  labs(title = paste("Correlation plot of the liver modules after", exp),
       x = "Eigengene score of the mouse 48h liver modules",
       y = paste("Eigengene score of the liver modules for", exp))+
  theme_minimal() +
      scale_color_manual(values = "green")
    
    print(p)
    

    # Store the plot in the plot_list
    plot_list[[length(plot_list) + 1]] <- p
    
  } else {
    message(paste("No data available for experiment:", exp))
  }
}
```
##Create correlation plots for the mouse 48h model + pear r score

```{r}
# Initialize plot list
plot_list <- list() 

# Loop through each experiment and create a correlation plot
unique_experiments <- unique(eg_scores_dili_sili_filtered$experiment_dili)

for (exp in unique_experiments) {
  # Subset the data for the current experiment
  subset_data <- eg_scores_dili_sili_filtered %>% 
    filter(experiment_dili == exp) %>%
    filter(experiment_sili == "mouse 48h")
  
  # Check if subset_data is non-empty
  if (nrow(subset_data) > 0) {
    # Identify the top 5 and bottom 5 modules based on eg_score
    top_5_dili <- subset_data %>% top_n(5, eg_score_dili)
    top_5_sili <- subset_data %>% top_n(5, eg_score_sili)
    bottom_5_dili <- subset_data %>% top_n(-5, eg_score_dili)
    bottom_5_sili <- subset_data %>% top_n(-5, eg_score_sili)
  
    # Label the top and bottom modules
    label_data1 <- bind_rows(top_5_dili, bottom_5_dili) %>%
      mutate(experiment_sili = case_when(
        experiment_sili == "uploadFile2_liverhepatocytesPHx48hr" ~ "rat_dataset",
        experiment_sili == "uploadFile1_NCSUPHxcaudate" ~ "mouse_dataset",
        TRUE ~ as.character(experiment_sili)
      ))

    label_data2 <- bind_rows(bottom_5_sili, top_5_sili) %>%
      mutate(experiment_sili = case_when(
        experiment_sili == "uploadFile2_liverhepatocytesPHx48hr" ~ "rat_dataset",
        experiment_sili == "uploadFile1_NCSUPHxcaudate" ~ "mouse_dataset",
        TRUE ~ as.character(experiment_sili)
      ))

    label_data <- rbind(label_data2, label_data1)
    
    # Calculate the correlation coefficient
    R_score <- cor(subset_data$eg_score_sili, subset_data$eg_score_dili)

    p <- ggplot(subset_data, aes(x = eg_score_sili, y = eg_score_dili, color = experiment_sili)) +
      geom_point() +
      geom_smooth(method = "lm", se = FALSE) + 
      geom_label_repel(data = label_data1, aes(label = module_number), 
                       box.padding = 0.1, color = "red", segment.color = "grey50") +
      geom_label_repel(data = label_data2, aes(label = module_number), 
                       box.padding = 0.8, color = "blue", segment.color = "grey50") +
      labs(title = paste("Correlation plot of the liver modules after", exp),
           x = "Eigengene score of the mouse 48h liver modules",
           y = paste("Eigengene score of the liver modules for", exp)) +
      theme_minimal() +
      scale_color_manual(values = "green") +
      annotate("text", x = Inf, y = Inf, label = paste("R =", round(R_score, 2)), hjust = 8.1, vjust = 2, size = 4, color = "black")
    
    print(p)
    
    # Store the plot in the plot_list
    plot_list[[length(plot_list) + 1]] <- p
    
  } else {
    message(paste("No data available for experiment:", exp))
  }
}

```




##select the top 10 compounds with the highest pearR score for mouse
```{r}
eg_scores_dili_filtered1 <- eg_scores_dili %>%
  filter(experiment == "PHALLOIDIN", conc == "1", time == "1") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered2 <- eg_scores_dili %>%
filter(experiment == "MONOCROTALINE", conc == "10", time== "8") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered3 <- eg_scores_dili %>%
filter(experiment == "DANAZOL", conc == "1000", time== "1") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered4 <- eg_scores_dili %>%
filter(experiment == "METHYLENE DIANILINE", conc == "100", time== "4") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered5 <- eg_scores_dili %>%
filter(experiment == "TRIMETHADIONE", conc == "500", time== "1") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered6 <- eg_scores_dili %>%
filter(experiment == "THIOACETAMIDE", conc == "45", time== "4") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered7 <- eg_scores_dili %>%
filter(experiment == "THIOACETAMIDE", conc == "45", time== "15") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))%>%
  mutate(experiment_dili = paste0(experiment_dili, "2"))

eg_scores_dili_filtered8 <- eg_scores_dili %>%
filter(experiment == "CARBON TETRACHLORIDE", conc == "300", time== "15") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))

eg_scores_dili_filtered9 <- eg_scores_dili %>%
filter(experiment == "THIOACETAMIDE", conc == "45", time== "29") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))%>%
  mutate(experiment_dili = paste0(experiment_dili, "3"))

eg_scores_dili_filtered10 <- eg_scores_dili %>%
filter(experiment == "MONOCROTALINE", conc == "10", time== "15") %>%
  merge(eg_scores_sili_filtered, by = "module", suffixes = c("_dili", "_sili"))%>%
  mutate(experiment_dili = paste0(experiment_dili, "2"))

eg_scores_dili_sili_filtered_mouse <- rbind(
  # eg_scores_dili_filtered1, eg_scores_dili_filtered2, eg_scores_dili_filtered3) 
  # eg_scores_dili_filtered4, eg_scores_dili_filtered5,
  eg_scores_dili_filtered6, eg_scores_dili_filtered7, eg_scores_dili_filtered8)
  # eg_scores_dili_filtered9, eg_scores_dili_filtered10)
```

```{r}
eg_scores_dili_sili_filtered_mouse <- eg_scores_dili_sili_filtered_mouse %>%
  mutate(experiment_sili = case_when(
    experiment_sili == "uploadFile1_NCSUPHxcaudate" ~ "rat 48h",
    experiment_sili == "uploadFile2_liverhepatocytesPHx48hr" ~ "mouse 48h",
    TRUE ~ experiment_sili  # Keep other values unchanged
  ))

```


##Create correlation plots for the rat 48h model
```{r}

# Initialize plot list
plot_list_mouse <- list()

# Loop through each experiment and create a correlation plot
unique_experiments_mouse <- unique(eg_scores_dili_sili_filtered_mouse$experiment_dili)

for (exp in unique_experiments_mouse) {
  # Subset the data for the current experiment
  subset_data_mouse <- eg_scores_dili_sili_filtered_mouse %>% filter(experiment_dili == exp)%>%
    filter(experiment_sili == "rat 48h")
  
  # Check if subset_data is non-empty
  if (nrow(subset_data_mouse) > 0) {
   
    # Identify the top 10 and bottom 10 modules based on eg_score
    top_5_dili_mouse <- subset_data_mouse %>% top_n(5, eg_score_dili)
    top_5_sili_mouse<- subset_data_mouse %>% top_n(5, eg_score_sili)
    bottom_5_dili_mouse <- subset_data_mouse %>% top_n(-5, eg_score_dili)
    bottom_5_sili_mouse <-subset_data_mouse %>% top_n(-5, eg_score_sili)
    
   # Label the top and bottom modules
    label_data_mouse1 <- bind_rows( bottom_5_dili_mouse, top_5_dili_mouse) %>%
      mutate(experiment_sili = case_when(
        experiment_sili == "uploadFile2_liverhepatocytesPHx48hr" ~ "rat_dataset",
        experiment_sili == "uploadFile1_NCSUPHxcaudate" ~ "mouse_dataset",
        TRUE ~ as.character(experiment_sili)
      ))
     
    
    label_data_mouse2 <- bind_rows( bottom_5_sili_mouse, top_5_sili_mouse) %>%
      mutate(experiment_sili = case_when(
        experiment_sili == "uploadFile2_liverhepatocytesPHx48hr" ~ "rat_dataset",
        experiment_sili == "uploadFile1_NCSUPHxcaudate" ~ "mouse_dataset",
        TRUE ~ as.character(experiment_sili)
      ))
    
    
    label_data <- bind_rows(top_5_dili, bottom_5_dili, top_5_sili, bottom_5_sili) %>%
      mutate(experiment_sili = case_when(
        experiment_sili == "uploadFile2_liverhepatocytesPHx48hr" ~ "rat_dataset",
        experiment_sili == "uploadFile1_NCSUPHxcaudate" ~ "mouse_dataset",
        TRUE ~ as.character(experiment_sili)
      ))
    
    label_data_mouse <- rbind(label_data_mouse2, label_data_mouse1)


    p <- ggplot(subset_data_mouse, aes(x = eg_score_sili, y = eg_score_dili, color = experiment_sili)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  geom_label_repel(data = label_data_mouse1, aes(label = module_number), 
                   box.padding = 1.5, color = "red", segment.color = "grey50") +
  geom_label_repel(data = label_data_mouse2, aes(label = module_number), 
                   box.padding = 0.2, color = "blue", segment.color = "grey50") +
  labs(title = paste("Correlation plot of the liver modules after", exp),
       x = "Eigengene score of the rat 48h liver modules",
       y = paste("Eigengene score for the liver modules for", exp))+
  theme_minimal() +
      scale_color_manual(values = "orange")
   
  
    print(p)

    # Store the plot in the plot_list
    plot_list_mouse[[length(plot_list_mouse) + 1]] <- p
    
  } else {
    message(paste("No data available for experiment:", exp))
  }
}

```

##Create correlation plots for the rat 48h model + pear r  score in plot
```{r}
# Initialize plot list
plot_list_mouse <- list()

# Loop through each experiment and create a correlation plot
unique_experiments_mouse <- unique(eg_scores_dili_sili_filtered_mouse$experiment_dili)

for (exp in unique_experiments_mouse) {
  # Subset the data for the current experiment
  subset_data_mouse <- eg_scores_dili_sili_filtered_mouse %>% 
    filter(experiment_dili == exp) %>%
    filter(experiment_sili == "rat 48h")
  
  # Check if subset_data is non-empty
  if (nrow(subset_data_mouse) > 0) {
    # Identify the top 5 and bottom 5 modules based on eg_score
    top_5_dili_mouse <- subset_data_mouse %>% top_n(5, eg_score_dili)
    top_5_sili_mouse <- subset_data_mouse %>% top_n(5, eg_score_sili)
    bottom_5_dili_mouse <- subset_data_mouse %>% top_n(-5, eg_score_dili)
    bottom_5_sili_mouse <- subset_data_mouse %>% top_n(-5, eg_score_sili)
    
    # Label the top and bottom modules
    label_data_mouse1 <- bind_rows(bottom_5_dili_mouse, top_5_dili_mouse) %>%
      mutate(experiment_sili = case_when(
        experiment_sili == "uploadFile2_liverhepatocytesPHx48hr" ~ "rat_dataset",
        experiment_sili == "uploadFile1_NCSUPHxcaudate" ~ "mouse_dataset",
        TRUE ~ as.character(experiment_sili)
      ))

    label_data_mouse2 <- bind_rows(bottom_5_sili_mouse, top_5_sili_mouse) %>%
      mutate(experiment_sili = case_when(
        experiment_sili == "uploadFile2_liverhepatocytesPHx48hr" ~ "rat_dataset",
        experiment_sili == "uploadFile1_NCSUPHxcaudate" ~ "mouse_dataset",
        TRUE ~ as.character(experiment_sili)
      ))

    label_data_mouse <- rbind(label_data_mouse2, label_data_mouse1)

    # Calculate the correlation coefficient
    R_score <- cor(subset_data_mouse$eg_score_sili, subset_data_mouse$eg_score_dili)

    p <- ggplot(subset_data_mouse, aes(x = eg_score_sili, y = eg_score_dili, color = experiment_sili)) +
      geom_point() +
      geom_smooth(method = "lm", se = FALSE) +
      geom_label_repel(data = label_data_mouse1, aes(label = module_number), 
                       box.padding = 1.5, color = "red", segment.color = "grey50") +
      geom_label_repel(data = label_data_mouse2, aes(label = module_number), 
                       box.padding = 0.2, color = "blue", segment.color = "grey50") +
      labs(title = paste("Correlation plot of the liver modules after", exp),
           x = "Eigengene score of the rat 48h liver modules",
           y = paste("Eigengene score for the liver modules for", exp)) +
      theme_minimal() +
      scale_color_manual(values = "orange") +
      annotate("text", x = Inf, y = Inf, label = paste("R =", round(R_score, 2)), hjust = 8.1, vjust = 2, size = 4, color = "black")
    
    # Store the plot in the plot_list
    plot_list_mouse[[length(plot_list_mouse) + 1]] <- p
    
    print(p)

    #
  }
}

```





```{r}
library(cowplot)

# Combine the plots into a grid
combined_plot_new1 <- plot_grid(
  plotlist = plot_list, nrow = 1, align = 'h', rel_widths = rep(1, length(plot_list))
)

combined_plot_new2 <- plot_grid(
  plotlist = plot_list_mouse, nrow = 1, align = 'h', rel_widths = rep(1, length(plot_list_mouse))
)

# Combine the two rows into one plot
combined_plot <- plot_grid(combined_plot_new1, combined_plot_new2, nrow = 2)

# Save the combined plot as a PNG file
ggsave("combined_correlation_plots_0107.png", combined_plot, width = 18, height = 14, units = "in")

```


##export the correlation plots
```{r}
library(patchwork)
combined_correlation_plot_rat_mice <- rbind(plot_list_mouse, plot_list)
combined_plot_new1 <- wrap_plots(plotlist = plot_list, nrow = 2)  # Adjust nrow based on the number of plots you want in each row
# Save the combined plot as a PNG file
ggsave("combined_correlation_plots.png", plot_list, width = 18, height = 14, units = "in")

```


