---
title: "BOO2024 - Hands-on workshop DEG analysis"
author: "Imke Bruns (MSc), Steven Kunnen (PhD), Giulia Callegaro (PhD), Hugo van Kessel (Msc)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: default
    highlight: kate
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

# Setup {.tabset}
```{r include=FALSE, echo=TRUE, message=FALSE}
rm(list = ls()); gc()
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
```


## Load packages
### Bioconductor
```{r}
# Check if BiocManager is available and install
if(!require("BiocManager", quietly = T)){install.packages("BiocManager")}; library(BiocManager)

# Check if BiocManager is available and install
if(!require("DESeq2", quietly = T)){BiocManager::install("DESeq2", force = TRUE)}; library(DESeq2)
```


### CRAN
```{r}
# Check if pacman is available and install
if(!require("pacman", quietly = T)){install.packages("pacman")}; library(pacman)

# use packman to install CRAN packages
p_load(tidyverse, ggpubr, data.table, ggfortify, stats)


```


## Set directories
```{r}
# input directory
if(!dir.exists("INPUT")){
  dir.create(path = file.path(getwd(), "INPUT"))
}
input_dir <- file.path(getwd(), "INPUT")

# output directory
if(!dir.exists("OUTPUT")){
  dir.create(path = file.path(getwd(), "OUTPUT"))
}
output_dir <- file.path(getwd(), "OUTPUT")

```


## Load functions
Functions can also be loaded from a source script, but for clarity we will include them in the script.
```{r}

# Function: tidy check deseq input  ----
tidy_check = function(countdata, metadata) {
  # We assume first column will contain the row names (tidy = T in DESeqDataSetFromMatrix)
  names_count = names(countdata)[-1]
  # Not obligatory for DEseq2, but we want row names in the first column
  names_meta = metadata[[1]]
  # Test if the names from the count and metadata are identical (same entries, same order)
  identical = identical(names_count, names_meta)
  # If not identical, do we have the same entries (ignores duplicates)
  setequal = identical || setequal(names_count, names_meta)
  # Test for duplicates (in theory, but very unlikely, this can also happen if identical is true)
  has_duplicates = anyDuplicated(names_meta) > 0
  # If the names are not identical, but we're looking at the same entries without duplicates,
  # then the problem is the order of the entries
  problem_is_order = !identical && setequal && !has_duplicates
  # List possibilities to test
  input = list(identical = identical, setequal = setequal, has_duplicates = has_duplicates, problem_is_order = problem_is_order)

  if(input$identical & input$setequal & !input$has_duplicates & !input$problem_is_order){
    return(TRUE)
  } else {
    print(input)
    return(FALSE)
  }
}

```



# Load data {.tabset}
Here you should load the data you obtained from the quality control (QC) workshop last week.

## Metadata
We name the object `metadata`, try to read the data into R yourself.
```{r}
# metadata = 
metadata<-  read_delim(file.path(input_dir,"20240517_metadata_processed.csv"))

metadata<-
  metadata %>%
  select(SAMPLE_ID, everything())

```


## Raw counts
We name the object `countdata_raw`, try to read the data into R yourself.
```{r}
# countdata_raw = 
 countdata_raw <- read_delim(file.path(input_dir, "20240517_countdata_raw_processed.csv"))

new_colnames <- c("GENE_ID", "PHx_0HR_R1", "PHx_0HR_R2", "PHx_0HR_R3", "PHx_48HR_R1", "PHx_48HR_R2", "PHx_48HR_R3")

colnames(countdata_raw) <-new_colnames
countdata_raw_average<- countdata_raw %>%
  mutate(PHx_0h = (PHx_0HR_R1 + PHx_0HR_R2 + PHx_0HR_R3) / 3)%>%
  mutate(PHx_48h = (PHx_48HR_R1 + PHx_48HR_R2 + PHx_48HR_R3) / 3)%>%
  select(GENE_ID, PHx_0h, PHx_48h)%>%
  na.omit(GENE_ID)

```


# Differential gene expression analysis based on the negative binomial distribution (DESeq2 package) {.tabset}

## Generate contrast
We first need to generate a contrast to tell the DESeq2 function which comparisons we want to make. Thus, here we match each treatment condition with a control condition. This is usually done per timepoint (we have a control condition for each of the timepoints). Examples: 
* treatmentA_1uM_4hr vs control_4hr  
* treatmentA_5uM_4hr vs control_4hr  
* treatmentA_1uM_8hr vs control_8hr  
* treatmentB_3uM_4hr vs control_4hr   


Questions:   
* What does the mean_id_treatment and mean_id_control column represent? hint: take a look at the metadata object   

### Set control condition
Check in your data what would be the control condition for your samples. The control condition can be DMSO/DMEM/MEDIUM/STARVMEDIUM/NaCl/etc. If there are multiple control conditions; create vector like this: c("DMSO", "DMEM")
```{r error=F,warning=F,message=F, echo=F}

# control_condition = 
"liver hepatocytes_PHx_0hr"

```


### Create the `contrast_table`
```{r error=F,warning=F,message=F, echo=F}
# We generate a contrast table which contains the mean_id comparison we aim the perform
contrast_control = metadata %>%
  filter( Time_point== "0hr") %>%
   distinct(MEAN_ID, Time_point, protocol, Organism)


contrast_treatment = metadata %>%
  filter( Time_point== "48hr") %>%
   distinct(MEAN_ID, Time_point, protocol, Organism)


contrast_table = left_join(x = contrast_treatment %>%
                             select(MEAN_ID_TREATMENT = MEAN_ID, Time_point, Organism, EXPERIMENT = protocol),
                           y = contrast_control %>%
                             select(MEAN_ID_CONTROL = MEAN_ID, Time_point, Organism,  CONTROL = protocol),
                           by = "Organism") %>%
  distinct()


contrast_table %>%
  select(MEAN_ID_TREATMENT, MEAN_ID_CONTROL) %>%
  as_tibble() %>%
  print(n = 1000)


```



## Run DESeq2

Questions:   
* What can we conclude for each treatment condition?   
* What is a differentially expressed gene?   
* Why do we use padj and log2FC as threshold to determine if a gene is differentially expressed   
* Give a definition of the following columns: baseMean, log2FoldChange, lfcSE, pvalue and padj   
   
   
```{r}
countdata_raw<- na.omit(countdata_raw)
tidy_check(countdata = countdata_raw, metadata = metadata)
```

```{r error=F,warning=F,message=F, echo=F}

# We set the thresholds to find differential expressed genes
padj_threshold <- 0.05
log2FC_threshold <- log2(1.5)

# We loop over the contrast table and select the corresponding data, make the DESeq object, cpm normalize, contrast the treatment with the control and print the ouput
deseq_results = tibble()


  if(nrow(metadata)==ncol(countdata_raw)-1){
    
    if(tidy_check(countdata = countdata_raw, metadata = metadata)){
      deseq_object = DESeqDataSetFromMatrix(countData =  as.data.frame(countdata_raw),
                                                 colData = metadata %>% mutate(MEAN_ID = as.factor(MEAN_ID)),
                                                 design = ~ MEAN_ID,
                                                 tidy = TRUE)
    }
    sizeFactors(deseq_object) = colSums(column_to_rownames(countdata_raw, var = "GENE_ID"))/1E6
    deseq_object = DESeq(deseq_object)

    for(i in 1:nrow(contrast_table)){
    temp = data.frame(
      results(deseq_object, contrast = c("MEAN_ID", contrast_table$MEAN_ID_TREATMENT[i], contrast_table$MEAN_ID_CONTROL[i]))) %>%
      rownames_to_column(var = "GENE_ID") %>%
      mutate(MEAN_ID_TREATMENT = contrast_table$MEAN_ID_TREATMENT[i], 
             MEAN_ID_CONTROL = contrast_table$MEAN_ID_CONTROL[i]) %>% tibble()
    
    
    up = temp %>% filter(padj < padj_threshold & log2FoldChange >= log2FC_threshold) %>% nrow()
    down = temp %>% filter(padj < padj_threshold & log2FoldChange <= -log2FC_threshold) %>% nrow()
    
    cat(
      paste("Treatment", contrast_table$MEAN_ID_TREATMENT[i], "vs. control", contrast_table$MEAN_ID_CONTROL[i]), 
      "\npadj threshold =", padj_threshold, "log2FC threshold =", log2FC_threshold, 
      "\nUp regulated DEGs =", up, "\nDown regulated DEGs =", down)
    cat("\n")
    cat("\n")
    
    deseq_results = deseq_results %>% bind_rows(temp)
  }
}

# We print the DESeq result table
{
  cat("\n")
  print("DESeq2 output table")
  cat("\n")
  deseq_results %>% str()  
}
```


# Inspect DESeq2 output {.tabset}

## Log2FC vs FC

Questions:   
* Log2FC vs FC, what is the difference? Take a look at the plots. Why do we use the log2FC and not the FC?   

```{r error=F,warning=F,message=F}
# We plot a density plot using the foldchange values of the treatment conditions
deseq_results %>% 
  select(log2FoldChange, MEAN_ID_TREATMENT) %>% 
  mutate(FoldChange = 2^log2FoldChange) %>% 
  ggplot() +
  geom_density(mapping = aes(x = FoldChange)) +
  facet_wrap(~MEAN_ID_TREATMENT) +
  theme_bw() + 
  labs(title = "Density plot of FoldChange values")

# We ZOOM on x = c(0,10) and plot a density plot using the foldchange values of the treatment conditions
deseq_results %>% 
  select(log2FoldChange, MEAN_ID_TREATMENT) %>% 
  mutate(FoldChange = 2^log2FoldChange) %>% 
  ggplot() +
  geom_density(mapping = aes(x = FoldChange)) +
  facet_wrap(~MEAN_ID_TREATMENT) +
  theme_bw() + 
  labs(title = "ZOOM on x = c(0,10) Density plot of FoldChange values") +
  coord_cartesian(xlim = c(0,10)) +
  scale_x_continuous(breaks = 0:10)

# We plot a density plot using the log2foldchange values of the treatment conditions
deseq_results %>% 
  select(log2FoldChange, MEAN_ID_TREATMENT) %>% 
  mutate(FoldChange = 2^log2FoldChange) %>% 
  ggplot() +
  geom_density(mapping = aes(x = log2FoldChange)) +
  facet_wrap(~MEAN_ID_TREATMENT) +
  theme_bw() + 
  labs(title = "Density plot of log2FoldChange values")

```



## Number of differentially expressed genes (DEGs)

Questions:   
* To inspect DEG count in your samples it is very quick to make a bar plot, but what are the drawbacks of this very simple DEG bar plot?   

```{r error=F,warning=F,message=F}
padj_threshold <- 0.05
log2FC_threshold <- log2(1.5)

deseq_results_sign_upregulated<- 
 deseq_results %>% 
  ##group_by(MEAN_ID_TREATMENT,.drop = F) %>% 
  filter(log2FoldChange > log2FC_threshold & padj < padj_threshold)

deseq_results_sign_downregulated<- 
 deseq_results %>% 
  ##group_by(MEAN_ID_TREATMENT,.drop = F) %>% 
  filter(log2FoldChange < -log2FC_threshold & padj < padj_threshold)

deseq_results_sign<- 
 deseq_results %>% 
  ##group_by(MEAN_ID_TREATMENT,.drop = F) %>% 
  filter(abs(log2FoldChange) > log2FC_threshold & padj < padj_threshold)
  
  ##count(MEAN_ID_TREATMENT) %>% 
  # ggplot(mapping = aes(x = MEAN_ID_TREATMENT,y = n)) + 
  # geom_bar(stat = "identity") +
  # theme_bw() +
  # theme(axis.text.x = element_text(angle = 90, size=4))
   
 deseq_results_sign %>%
   nrow()
 
 total_sign_genes<- rbind(deseq_results_sign_upregulated, deseq_results_sign_downregulated, deseq_results_sign)
 

```

### Volcano plot

Questions:   
* Why do we want to inspect a volcano plot? What conclusion can we make for each treatment condition?   

```{r error=F,warning=F,message=F}
padj_threshold <- 0.05
log2FC_threshold <- log2(1.5)

# We assign a status to every gene, either significant or not, based on the thresholds
red = deseq_results %>% 
  mutate(class = if_else(condition = abs(log2FoldChange) > log2FC_threshold & padj < padj_threshold,
                         true = "significant",
                         false = "not significant")) %>% 
  drop_na()

# We generate a scatter plot coloring the significant and not significant genes per treatment condition
plot <- ggplot(data = red, mapping = aes(x = log2FoldChange, y = -log10(padj), color = class)) + 
  facet_wrap(~MEAN_ID_TREATMENT) +
  geom_point(size = 0.5) +
  theme_bw()

# We print the output to generate the plot
plot

```


## MA plot

Questions:   
* What conclusions can you make for the red data points in the the MA plot?   
* What can you conclude for genes with baseMean < 2.5 and log2FC > 2.5? Are these log2FC values biologically significant?

```{r error=F,warning=F,message=F}
# Assign baseMean and log2FC thresholds
baseMean_threshold <- 2.5
log2FC_threshold <- 2.5
padj_threshold <- 0.05

# We generate the plot and print the output
ggplot(data = deseq_results, aes(x = baseMean, y = log2FoldChange)) + 
  geom_point(size=0.5) +
  geom_point(data = deseq_results %>% filter(padj < padj_threshold), color = "red", size=0.5) +
  facet_wrap(~MEAN_ID_TREATMENT) +
  scale_x_continuous(limits = c(0,10)) +
  theme_bw() +
  labs(title = "MA plot with significant genes in red")
```


#### Principal component analysis on DESeq output using the log2FC values

Questions:   
* What can you conclude from the PCA on the log2FC of the treatment conditions? 


```{r error=F,warning=F,message=F}

# We transpose prepare the data frame for pca analysis
log2fc_results <- deseq_results %>%
  select(GENE_ID, log2FoldChange, MEAN_ID_TREATMENT) %>%
  pivot_wider(names_from = MEAN_ID_TREATMENT, values_from = log2FoldChange)

# We print the genes with NA as log2FC result
print("NA genes dropped from PCA analysis")
log2fc_results[!!rowSums(is.na(log2fc_results)),]

# We further wrangle the data frame for pca analysis
pca_data <- log2fc_results %>% 
  drop_na() %>% 
  pivot_longer(cols = where(is.numeric), names_to = "MEAN_ID") %>% 
  pivot_wider(values_from = value, names_from = GENE_ID) %>%
  left_join(metadata, by="MEAN_ID")

# We perform pca analysis on numerical columns
pca_object = prcomp(pca_data %>% select(log2fc_results$GENE_ID), center = F, scale. = F)

# We print the output
  autoplot(object = pca_object, data = pca_data, colour = "Time_point", size="protocol") + 
    theme_bw()
```

# Rat orthology database RGD
 
```{r include=TRUE, echo=TRUE, message=FALSE}
if(!require("dplyr", quietly = T)){BiocManager::install("dplyr", force=TRUE)};
library(dplyr)

rat_rgd <- fread(input = "https://download.rgd.mcw.edu/data_release/RAT/ORTHOLOGS_RAT.txt",
                 stringsAsFactors = FALSE, 
                 fill = TRUE, 
                 skip = 15, 
                 header = "auto", 
                 na.strings = c(NA_character_, ""))
 
rat_rgd <- rat_rgd[, !duplicated(colnames(rat_rgd)), with = F] %>% data.frame() %>% 
  select(RAT_GENE_SYMBOL,
         RAT_GENE_NCBI_GENE_ID,
         MOUSE_GENE_SYMBOL,
         MOUSE_GENE_NCBI_GENE_ID,
         MOUSE_GENE_ENSEMBL_GENE_ID) %>% tibble()

deseq_results_mouse_to_rat_gene <- deseq_results%>% 
  tidyr::separate(col= "GENE_ID", into = c("GENE_ID", "VERSION"), sep ="[.]")%>%
  left_join(rat_rgd %>% select(RAT_GENE_NCBI_GENE_ID, "GENE_ID" = MOUSE_GENE_ENSEMBL_GENE_ID), by = "GENE_ID")


##deseq_results_dataset2_rat <- deseq_results %>%
  ##mutate(MOUSE_GENE_ENSEMBL_GENE_ID = gsub("\\..*", "", GENE_ID))


# # deseq_results_mouse_to_rat_gene <- deseq_results_dataset2_rat %>% 
#   left_join(rat_rgd %>% select(RAT_GENE_ENSEMBL_GENE_ID, MOUSE_GENE_ENSEMBL_GENE_ID), by = "MOUSE_GENE_ENSEMBL_GENE_ID") %>%
#   select(-GENE_ID, -MOUSE_GENE_ENSEMBL_GENE_ID)%>%
#   select(RAT_GENE_ENSEMBL_GENE_ID, everything())

```


# Create expression data
```{r}

##deseq_results_mouse_to_rat_gene$experiment <- deseq_results_dataset2_rat$MEAN_ID_TREATMENT 

expression_long<- deseq_results_mouse_to_rat_gene  %>% 
  dplyr::select(experiment = MEAN_ID_TREATMENT,
                gene_id = RAT_GENE_NCBI_GENE_ID,
                log2fc = log2FoldChange,
                pvalue,
                padj)

expression_long<- expression_long %>%
  mutate( gene_id = as.integer(gene_id))

##expression_long$time <- as.character(expression_long$time)

# add columns needed for TXG-MAPr

expression_long_TXG <- left_join(expression_long, metadata %>%
  dplyr::select("experiment" = "MEAN_ID", "time" = "Time_point")%>%
    unique(),
  by = c("experiment"))
  
  expression_long_TXG <- expression_long_TXG %>%
  mutate(time = 2, conc = 1)

  
```



```{r}
dist(x)
```


# Save output
```{r}
write_rds(deseq_results, file.path(output_dir, paste0(gsub( "-", "", Sys.Date()), "_DESeq2_results_unformatted_dataset2.rds"))) 
write.table(expression_long_TXG, file.path(output_dir, paste0(gsub( "-", "", Sys.Date()), "_expression_long_TXG_upload3.txt")), sep = "\t", row.names = F)
```


