---
title: "BOO2024 - Hands-on workshop DEG analysis"
author: "Imke Bruns (MSc), Steven Kunnen (PhD), Giulia Callegaro (PhD), Hugo van Kessel (Msc)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: default
    highlight: kate
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

# Setup {.tabset}
```{r include=FALSE, echo=TRUE, message=FALSE}
rm(list = ls()); gc()
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
```


## Load packages
### CRAN
```{r}
# Check if pacman is available and install
if(!require("pacman", quietly = T)){install.packages("pacman")}; library(pacman)

# use packman to install CRAN packages
p_load(tidyverse, data.table, enrichR, fgsea)

```


## Set directories
```{r}
# input directory
if(!dir.exists("INPUT")){
  dir.create(path = file.path(getwd(), "INPUT"))
}
input_dir <- file.path(getwd(), "INPUT")

# output directory
if(!dir.exists("OUTPUT")){
  dir.create(path = file.path(getwd(), "OUTPUT"))
}
output_dir <- file.path(getwd(), "OUTPUT")

```


# Load data {.tabset}
Here you should load the data you obtained from the quality control (QC) workshop last week.

## Metadata
We name the object `metadata`, try to read the data into R yourself.
```{r}
 metadata = read_delim(file.path(input_dir,"20240517_metadata_processed.csv"))

```


## DESeq2 output
We name the object `deseq_results`, try to read the data into R yourself.
```{r}
deseq_results = readRDS(file.path(output_dir,"20240522_DESeq2_results_unformatted.rds"))

```


```{r}
url_link = "https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/Mammalia/"
download.file(paste0(url_link, "Mus_musculus.gene_info.gz"), destfile = paste0(input_dir, "/Mus_musculus.gene_info.gz"), method = "auto")
NCBI_gene <- fread(paste0(input_dir, "/Mus_musculus.gene_info.gz")) |>
  mutate(dbXrefs = str_replace_all(dbXrefs, ":MGI:", ":"))

# https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/Mammalia/Mus_musculus.gene_info.gz
NCBI_gene <- NCBI_gene|>
  mutate(mgi_id = str_extract(dbXrefs, pattern = "MGI[:]\\d+"),
         # hgnc_id = str_extract(dbXrefs, pattern = "HGNC[:]\\d+"),
         ensembl_id = str_extract(dbXrefs, pattern = "Ensembl[:]ENSMUSG\\d+"),
         alliancegenome_id = str_extract(dbXrefs, pattern = "AllianceGenome[:]\\d+")) |>
  mutate(
    mgi_id = str_replace(mgi_id, pattern = "MGI[:]", ""),
         # hgnc_id = str_replace(hgnc_id, pattern = "HGNC[:]", ""),
         ensembl_id = str_replace(ensembl_id, pattern = "Ensembl[:]", ""),
         alliancegenome_id = str_replace(alliancegenome_id, pattern = "AllianceGenome[:]", ""),
         # Symbol = paste0("id_", Symbol),
         # GeneID = paste0("id_", GeneID),
         user_date = as.integer(str_remove_all(Sys.Date(), "-"))) |>
  dplyr::select("entrez_id" = GeneID, "gene_symbol" = Symbol, ensembl_id, mgi_id, alliancegenome_id, "modification_date" = Modification_date,user_date)


```


```{r}
if(!require("dplyr", quietly = T)){BiocManager::install("dplyr", force=TRUE)};
library(dplyr)

NCBI_gene_filtered <- NCBI_gene %>%
  select(gene_symbol, ensembl_id)


deseq_results_2 <- deseq_results %>%
  mutate(ensembl_id = gsub("\\..*", "", GENE_ID))


deseq_results_gene_symbol <- deseq_results_2  %>% 
  left_join(NCBI_gene_filtered %>% select(gene_symbol, ensembl_id), by = "ensembl_id")

deseq_results_enrichR <- deseq_results_gene_symbol %>%
  select(-GENE_ID)%>%
  select(gene_symbol, ensembl_id, everything())


```



# Gene set enrichment analysis {.tabset}

## Load gmt files
Gmt files can be downloaded from the msigdb database (https://www.gsea-msigdb.org/gsea/msigdb/). Load the pathway data of interest here using the `gmtPathways` function
```{r}
gmt_wikipathways = gmtPathways(file.path(
  input_dir,
  "GSEA_PATHWAYS",
  "wikipathways.v2023.2.Hs.symbols.gmt"
))

gmt_reactome = gmtPathways(file.path(
  input_dir,
  "GSEA_PATHWAYS",
  "reactome.v2023.2.Hs.symbols.gmt"
))

gmt_hallmarks = gmtPathways(file.path(
  input_dir,
  "GSEA_PATHWAYS",
  "hallmarks.v2023.2.Hs.symbols.gmt"
))

combined_gmt <- c(gmt_wikipathways,  gmt_reactome)

```

##combine rat and mouse data
```{r}
mouse_data<- deseq_results_enrichR%>%
  select(MEAN_ID_TREATMENT, gene_symbol, stat,)
rat_data <- read.delim(file.path(output_dir,"NCSU_PHx_DEG_Data_all_symbol.txt"))
rat_data<- rat_data%>% 
  filter(experiment_id %in% c('NCSU.PHx_Liver_PHx_Hepatectomy_0d_caudate',
                              'NCSU.PHx_Liver_PHx_Hepatectomy_0.042d_caudate',
                              'NCSU.PHx_Liver_PHx_Hepatectomy_0.125d_caudate',
                              'NCSU.PHx_Liver_PHx_Hepatectomy_0.25d_caudate',
                           'NCSU.PHx_Liver_PHx_Hepatectomy_0.5d_caudate',	
                          'NCSU.PHx_Liver_PHx_Hepatectomy_0.75d_caudate',				
'NCSU.PHx_Liver_PHx_Hepatectomy_1d_caudate',			
'NCSU.PHx_Liver_PHx_Hepatectomy_2d_caudate',		
'NCSU.PHx_Liver_PHx_Hepatectomy_3d_caudate',				
'NCSU.PHx_Liver_PHx_Hepatectomy_5d_caudate',			
'NCSU.PHx_Liver_PHx_Hepatectomy_10d_caudate',				
'NCSU.PHx_Liver_PHx_Hepatectomy_15d_caudate'
))%>%
  rename(MEAN_ID_TREATMENT = experiment_id, stat = t)

rat_data<- rat_data %>%
  select(gene_symbol, MEAN_ID_TREATMENT, stat)

combined_rat_mouse <- rbind(rat_data, mouse_data)
```



## Run fgsea

Questions:   
* Explain how fgsea determines which pathways are enriched. What is the difference between an Overrepresentation Analysis (ORA) and a Gene Set Enrichment Analysis (GSEA)? 
* We need to input a ranked gene list, we rank the genes based on their Wald statistic. What is the advantage of using the Wald statistic instead of for example the log2FoldChange? 

  
```{r}

gsea <- combined_rat_mouse %>%
  nest_by(MEAN_ID_TREATMENT) %>%
  mutate(ranked_genes = list(
    data %>% arrange(stat) %>% select(gene_symbol, stat) %>% deframe()
  )) %>%
  mutate(fgsea_output = list(fgsea(pathways = combined_gmt,
                                   stats = ranked_genes)))


# print the output
for (i in gsea$MEAN_ID_TREATMENT) {
  print(paste0("Top 10 significant pathways for ", i))
  print(
    gsea %>%
      filter(MEAN_ID_TREATMENT == i) %>%
      select(MEAN_ID_TREATMENT, fgsea_output) %>%
      unnest(cols = c(fgsea_output)) %>%
      select(MEAN_ID_TREATMENT, pathway, padj, NES) %>%
      arrange(desc(abs(NES))) %>%
      ungroup() %>%
      slice(1:30)
  )
}

```


```{r}
gsea_results <- gsea %>%
  unnest(cols = c(fgsea_output)) %>%
  select(MEAN_ID_TREATMENT, pathway, NES) %>%
  ungroup()


get_top_and_bottom_nes <- function(gsea_results, n = 5) {
  # # Add absolute NES column
  # gsea_results <- gsea_results %>%
  #   mutate(abs_NES = abs(NES))
  
  # Sort by NES within each condition
  sorted_results <- gsea_results %>%
    arrange(MEAN_ID_TREATMENT, desc(NES))
  
  # Select top 5 highest NES and bottom 5 lowest NES pathways per condition
  top_and_bottom_nes <- sorted_results %>%
    group_by(MEAN_ID_TREATMENT) %>%
    slice(c(1:n, n() - (0:(n-1)))) %>%  # Select top n and bottom n rows per group
    ungroup() %>%
    arrange(MEAN_ID_TREATMENT, desc(NES)) 
  
  return(top_and_bottom_nes)
}

# Usage example:
combined_results <- get_top_and_bottom_nes(gsea_results, n = 5)
```

##rename the MEAN_ID_TREATMENT
```{r}
GSEA_results_rat <- combined_results %>%
  filter(!MEAN_ID_TREATMENT  == "liver hepatocytes_PHx_48hr")

GSEA_results_mouse<- combined_results %>%
  filter(MEAN_ID_TREATMENT  == "liver hepatocytes_PHx_48hr")
```

##rename experiment_id
```{r}
transform_experiment_id <- function(MEAN_ID_TREATMENT) {
  # Extract the time point using regular expressions
  time_point <- str_extract(MEAN_ID_TREATMENT, "\\d+\\.?\\d*d")
  # Construct the new experiment_id
  new_experiment_id <- paste("NCSU.PHx", time_point, "rat", sep = "_")
  return(new_experiment_id)
}

GSEA_results_rat <- GSEA_results_rat %>%
  mutate(MEAN_ID_TREATMENT = sapply(MEAN_ID_TREATMENT, transform_experiment_id))

transform_experiment_id_mouse <- function(MEAN_ID_TREATMENT) {
  # Extract the time point using regular expressions
  time_point <- str_extract(MEAN_ID_TREATMENT, "\\d+(\\.\\d+)?d|\\d+h")
  # Construct the new experiment_id
  new_experiment_id <- paste("PHx", time_point, "mouse", sep = "_")
  return(new_experiment_id)
}

GSEA_results_mouse <- GSEA_results_mouse %>%
  mutate(MEAN_ID_TREATMENT = sapply(MEAN_ID_TREATMENT, transform_experiment_id_mouse))

GSEA_results_combined<- rbind(GSEA_results_mouse, GSEA_results_rat)

```

```{r}

desired_order <- c(
  "PHx_48h_mouse",
  "NCSU.PHx_0d_rat",
  "NCSU.PHx_0.042d_rat",
  "NCSU.PHx_0.125d_rat",
  "NCSU.PHx_0.25d_rat",
  "NCSU.PHx_0.5d_rat",
  "NCSU.PHx_0.75d_rat",
  "NCSU.PHx_1d_rat",
  "NCSU.PHx_2d_rat",
  "NCSU.PHx_3d_rat",
  "NCSU.PHx_5d_rat",
  "NCSU.PHx_10d_rat",
  "NCSU.PHx_15d_rat"
)

GSEA_results_combined$MEAN_ID_TREATMENT <- factor(GSEA_results_combined$MEAN_ID_TREATMENT, levels = desired_order)

```

##select the nes values for all pathways and models
```{r}
# combined_results_pathways contains all the unique pathways you want to select
combined_results_pathways <- GSEA_results_combined$pathway

# Initialize an empty data frame to store combined results
combined_filtered_results_rat_mice_2 <- data.frame()

# Extract fgsea_output and MEAN_ID_TREATMENT from gsea tibble
fgsea_output_list <- gsea$fgsea_output
mean_id_treatment_list <- unique(gsea$MEAN_ID_TREATMENT)

# Loop through each list in fgsea_output_list
for (i in seq_along(fgsea_output_list)) {
  # Extract the current list and corresponding MEAN_ID_TREATMENT
  current_list <- fgsea_output_list[[i]]
  mean_id_treatment <- mean_id_treatment_list[[i]]
  
  # Convert to data frame if necessary
  current_df <- as.data.frame(current_list)
  
  # Append MEAN_ID_TREATMENT to the current data frame
  current_df$MEAN_ID_TREATMENT <- mean_id_treatment
  
  # Filter the current data frame for the pathways in combined_results_pathways
  filtered_df <- current_df %>%
    filter(pathway %in% combined_results_pathways) %>%
    select(pathway, NES, MEAN_ID_TREATMENT)
  
  # Append the filtered data frame to the combined data frame
  combined_filtered_results_rat_mice_2 <- rbind(combined_filtered_results_rat_mice_2, filtered_df)
}
```

```{r}
GSEA_results_rat <- combined_results %>%
  filter(!MEAN_ID_TREATMENT  == "liver hepatocytes_PHx_48hr")

GSEA_results_mouse<- combined_results %>%
  filter(MEAN_ID_TREATMENT  == "liver hepatocytes_PHx_48hr")
```



```{r}

combined_filtered_results_rat_2<- combined_filtered_results_rat_mice_2%>%
  filter(!MEAN_ID_TREATMENT  == "liver hepatocytes_PHx_48hr")
  
combined_filtered_results_mouse_2<-combined_filtered_results_rat_mice_2%>%
  filter(MEAN_ID_TREATMENT  == "liver hepatocytes_PHx_48hr")


transform_experiment_id <- function(MEAN_ID_TREATMENT) {
  # Extract the time point using regular expressions
  time_point <- str_extract(MEAN_ID_TREATMENT, "\\d+\\.?\\d*d")
  # Construct the new experiment_id
  new_experiment_id <- paste("NCSU.PHx", time_point, "rat", sep = "_")
  return(new_experiment_id)
}

GSEA_results_rat <- combined_filtered_results_rat_2 %>%
  mutate(MEAN_ID_TREATMENT = sapply(MEAN_ID_TREATMENT, transform_experiment_id))

transform_experiment_id_mouse <- function(MEAN_ID_TREATMENT) {
  # Extract the time point using regular expressions
  time_point <- str_extract(MEAN_ID_TREATMENT, "\\d+(\\.\\d+)?d|\\d+h")
  # Construct the new experiment_id
  new_experiment_id <- paste("PHx", time_point, "mouse", sep = "_")
  return(new_experiment_id)
}

GSEA_results_mouse <- combined_filtered_results_mouse_2 %>%
  mutate(MEAN_ID_TREATMENT = sapply(MEAN_ID_TREATMENT, transform_experiment_id_mouse))

GSEA_results_combined_2<- rbind(GSEA_results_mouse, GSEA_results_rat)

desired_order <- c(
  "PHx_48h_mouse",
  "NCSU.PHx_0d_rat",
  "NCSU.PHx_0.042d_rat",
  "NCSU.PHx_0.125d_rat",
  "NCSU.PHx_0.25d_rat",
  "NCSU.PHx_0.5d_rat",
  "NCSU.PHx_0.75d_rat",
  "NCSU.PHx_1d_rat",
  "NCSU.PHx_2d_rat",
  "NCSU.PHx_3d_rat",
  "NCSU.PHx_5d_rat",
  "NCSU.PHx_10d_rat",
  "NCSU.PHx_15d_rat"
)

GSEA_results_combined_2$MEAN_ID_TREATMENT <- factor(GSEA_results_combined_2$MEAN_ID_TREATMENT, levels = desired_order)

```

##Make the heatmap
```{r}
library(tidyverse)
library(reshape2)
library(ggplot2)
library(plotly)

# Create the heatmap
heatmap_GSEA_ratmice <- ggplot(GSEA_results_combined_2, aes(x = MEAN_ID_TREATMENT, y = pathway, fill = NES)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  labs(title = "5 most up and down regulated pathways based on NES score from GSEA",
       x = "Experiment ID", y = "Pathway", fill = "NES") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 6, color = ifelse(levels(GSEA_results_combined_2$MEAN_ID_TREATMENT) == "PHx_48h_mouse", "green", "orange")),
    axis.text.y = element_text(size = 5),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 15)
  )

print(heatmap_GSEA_ratmice)
```

## export the heatmap
```{r}
library(reshape2)

ggsave("Heatmap_GSEA_2806_final.png", plot = heatmap_GSEA_ratmice, width =12, height = 6)
```



