---
title: "BOO2024 - Hands-on workshop DEG analysis"
author: "Imke Bruns (MSc), Steven Kunnen (PhD), Giulia Callegaro (PhD), Hugo van Kessel (Msc)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: default
    highlight: kate
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

# Setup {.tabset}
```{r include=FALSE, echo=TRUE, message=FALSE}
rm(list = ls()); gc()
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
```


## Load packages
### CRAN
```{r}
# Check if pacman is available and install
if(!require("pacman", quietly = T)){install.packages("pacman")}; library(pacman)

# use packman to install CRAN packages
p_load(tidyverse, data.table, enrichR, fgsea)

```


## Set directories
```{r}
# input directory
if(!dir.exists("INPUT")){
  dir.create(path = file.path(getwd(), "INPUT"))
}
input_dir <- file.path(getwd(), "INPUT")

# output directory
if(!dir.exists("OUTPUT")){
  dir.create(path = file.path(getwd(), "OUTPUT"))
}
output_dir <- file.path(getwd(), "OUTPUT")

```


## Load functions
Functions can also be loaded from a source script, but for clarity we will include them in the script.
```{r}


runEnrichR = function(genes) {
  enrichR = enrichr(
    genes = genes,
    databases = c(
      "BioCarta_2016",
      "HumanCyc_2016",
      "KEGG_2021_Human",
      "Reactome_2022",
      "WikiPathway_2021_Human",
      "InterPro_Domains_2019",
      "GO_Molecular_Function_2023",
      "GO_Cellular_Component_2023",
      "GO_Biological_Process_2023"
    )
  ) 
  
  enrichR = rbindlist(enrichR, idcol = T)
  names(enrichR)[1:2] = c("database", "source")
  return(enrichR)
}

```



# Load data {.tabset}
Here you should load the data you obtained from the quality control (QC) workshop last week.

## Metadata
We name the object `metadata`, try to read the data into R yourself.
```{r}
# metadata = 
metadata<-  read_delim(file.path(input_dir,"20240517_metadata_processed.csv"))

metadata<-
  metadata %>%
  select(SAMPLE_ID, everything())

```


## DESeq2 output
We name the object `deseq_results`, try to read the data into R yourself.
```{r}
# deseq_results = 
deseq_results <-  readRDS(file.path(output_dir,"20240522_DESeq2_results_unformatted.rds"))
```


# Overrepresentation analysis
Questions:   
* Which cellular processes are activated your cell type of interest after compound exposure?

```{r}
url_link = "https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/Mammalia/"
download.file(paste0(url_link, "Mus_musculus.gene_info.gz"), destfile = paste0(input_dir, "/Mus_musculus.gene_info.gz"), method = "auto")
NCBI_gene <- fread(paste0(input_dir, "/Mus_musculus.gene_info.gz")) |>
  mutate(dbXrefs = str_replace_all(dbXrefs, ":MGI:", ":"))

# https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/Mammalia/Mus_musculus.gene_info.gz
NCBI_gene <- NCBI_gene|>
  mutate(mgi_id = str_extract(dbXrefs, pattern = "MGI[:]\\d+"),
         # hgnc_id = str_extract(dbXrefs, pattern = "HGNC[:]\\d+"),
         ensembl_id = str_extract(dbXrefs, pattern = "Ensembl[:]ENSMUSG\\d+"),
         alliancegenome_id = str_extract(dbXrefs, pattern = "AllianceGenome[:]\\d+")) |>
  mutate(
    mgi_id = str_replace(mgi_id, pattern = "MGI[:]", ""),
         # hgnc_id = str_replace(hgnc_id, pattern = "HGNC[:]", ""),
         ensembl_id = str_replace(ensembl_id, pattern = "Ensembl[:]", ""),
         alliancegenome_id = str_replace(alliancegenome_id, pattern = "AllianceGenome[:]", ""),
         # Symbol = paste0("id_", Symbol),
         # GeneID = paste0("id_", GeneID),
         user_date = as.integer(str_remove_all(Sys.Date(), "-"))) |>
  dplyr::select("entrez_id" = GeneID, "gene_symbol" = Symbol, ensembl_id, mgi_id, alliancegenome_id, "modification_date" = Modification_date,user_date)


```


```{r}
if(!require("dplyr", quietly = T)){BiocManager::install("dplyr", force=TRUE)};
library(dplyr)

NCBI_gene_filtered <- NCBI_gene %>%
  select(gene_symbol, ensembl_id)


deseq_results_2 <- deseq_results %>%
  mutate(ensembl_id = gsub("\\..*", "", GENE_ID))


deseq_results_gene_symbol <- deseq_results_2  %>% 
  left_join(NCBI_gene_filtered %>% select(gene_symbol, ensembl_id), by = "ensembl_id")

deseq_results_enrichR <- deseq_results_gene_symbol %>%
  select(-GENE_ID)%>%
  select(gene_symbol, ensembl_id, everything())


```


##ORA for  treatment
```{r error=F,warning=F,message=F}

# We set the threshold for significant genes
padj_threshold <- 0.05

# We perform ORA analysis using EnrichR
ora <- deseq_results_enrichR %>%
  filter(padj < padj_threshold) %>%
  nest_by(MEAN_ID_TREATMENT) %>%
  mutate(significant_genes = list(data %>%
                                    arrange(padj) %>%
                                    # slice(1:50) %>%
                                    pull(gene_symbol))) %>%
  mutate(ora = list(
    runEnrichR(genes = significant_genes) %>% filter(Adjusted.P.value < 0.05)
  ))

# print the output
for (i in ora$MEAN_ID_TREATMENT) {
  print(paste0("Top 10 significant ORA for ", i))
  print(
    ora %>%
      filter(MEAN_ID_TREATMENT == i) %>%
      select(MEAN_ID_TREATMENT, ora) %>%
      unnest(cols = c(ora)) %>%
      select(MEAN_ID_TREATMENT, database, source, Adjusted.P.value, Genes) %>%
      arrange(Adjusted.P.value) %>%
      ungroup()%>%
      slice(1:20)
  )
}

```


```{r}
ORA_results <- ora %>%
  unnest(cols = c(ora)) %>%
  select(MEAN_ID_TREATMENT, source, Adjusted.P.value) %>%
  ungroup()

# Function to select top 5 highest NES and bottom 5 lowest NES pathways per condition
get_top_and_bottom_Adjusted.P.value <- function(ORA_results, n = 5) {
  # Add absolute Adjusted.P.value column
  ORA_results <- ORA_results %>%
    mutate(abs_Adjusted.P.value = abs(Adjusted.P.value))
  
  # Sort by Adjusted.P.value within each condition
  sorted_results <- ORA_results %>%
    arrange(MEAN_ID_TREATMENT, desc(Adjusted.P.value))
  
  # Select top 5 highest Adjusted.P.value and bottom 5 lowest Adjusted.P.value pathways per condition
  top_and_bottom_Adjusted.P.value <- sorted_results %>%
    group_by(MEAN_ID_TREATMENT) %>%
    slice(c(1:n, n() - (0:(n-1)))) %>%  # Select top n and bottom n rows per group
    ungroup() %>%
    arrange(MEAN_ID_TREATMENT, desc(Adjusted.P.value)) %>%  # Optional: reorder for clarity
    select(-abs_Adjusted.P.value)  # Remove abs_Adjusted.P.value column if not needed
  
  return(top_and_bottom_Adjusted.P.value)
}

# Usage example:
combined_results_ora <- get_top_and_bottom_Adjusted.P.value(ORA_results, n = 5)

write.table(combined_results_ora, file.path(output_dir, paste0(gsub( "-", "", Sys.Date()), "_ORA_results_mouse.txt")), sep = "\t", row.names = F)

```


##ORA for control
```{r error=F,warning=F,message=F}

# We set the threshold for significant genes
padj_threshold <- 0.05

# We perform ORA analysis using EnrichR
ora <- deseq_results_enrichR %>%
  filter(padj < padj_threshold) %>%
  nest_by(MEAN_ID_TREATMENT) %>%
  mutate(significant_genes = list(data %>%
                                    arrange(padj) %>%
                                    # slice(1:50) %>%
                                    pull(gene_symbol))) %>%
  mutate(ora = list(
    runEnrichR(genes = significant_genes) %>% filter(Adjusted.P.value < 0.05)
  ))

# print the output
for (i in ora$MEAN_ID_TREATMENT) {
  print(paste0("Top 10 significant ORA for ", i))
  print(
    ora %>%
      filter(MEAN_ID_TREATMENT == i) %>%
      select(MEAN_ID_TREATMENT, ora) %>%
      unnest(cols = c(ora)) %>%
      select(MEAN_ID_TREATMENT, database, source, Adjusted.P.value, Genes) %>%
      arrange(Adjusted.P.value) %>%
      ungroup()%>%
      slice(1:20)
  )
}

str(ora$significant_genes)

```


```{r}
ORA_results <- ora %>%
  unnest(cols = c(ora)) %>%
  select(MEAN_ID_TREATMENT, source, Adjusted.P.value) %>%
  ungroup()

# Function to select top 5 highest NES and bottom 5 lowest NES pathways per condition
get_top_and_bottom_Adjusted.P.value <- function(ORA_results, n = 5) {
  # Add absolute Adjusted.P.value column
  ORA_results <- ORA_results %>%
    mutate(abs_Adjusted.P.value = abs(Adjusted.P.value))
  
  # Sort by Adjusted.P.value within each condition
  sorted_results <- ORA_results %>%
    arrange(MEAN_ID_TREATMENT, desc(Adjusted.P.value))
  
  # Select top 5 highest Adjusted.P.value and bottom 5 lowest Adjusted.P.value pathways per condition
  top_and_bottom_Adjusted.P.value <- sorted_results %>%
    group_by(MEAN_ID_TREATMENT) %>%
    slice(c(1:n, n() - (0:(n-1)))) %>%  # Select top n and bottom n rows per group
    ungroup() %>%
    arrange(MEAN_ID_TREATMENT, desc(Adjusted.P.value)) %>%  # Optional: reorder for clarity
    select(-abs_Adjusted.P.value)  # Remove abs_Adjusted.P.value column if not needed
  
  return(top_and_bottom_Adjusted.P.value)
}

# Usage example:
combined_results_ora <- get_top_and_bottom_Adjusted.P.value(ORA_results, n = 5)

write.table(combined_results_ora, file.path(output_dir, paste0(gsub( "-", "", Sys.Date()), "_ORA_results_mouse.txt")), sep = "\t", row.names = F)

```
